\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{amsmath,amssymb,amsfonts,bm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\geometry{margin=25mm}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\title{Problem Formulation and Algorithm Candidates for Contextual-Bandit Ad Filtering in an In-House DSP}
\author{Matching Algorithm Research Note}
\date{\today}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\calA}{\mathcal{A}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calD}{\mathcal{D}}

\begin{document}
\maketitle

\begin{abstract}
This note reformulates the post-purchase advertising problem as a contextual-bandit
ad filtering problem for an in-house demand-side platform (DSP).  The objective is not
yet to solve real-time bidding end-to-end, but to construct a principled first layer
that selects a small set of promising ads from a large candidate pool under exploration,
cold-start, delayed feedback, and auction-induced selection bias.  We define the decision
process, reward, logging requirements, offline evaluation protocol, and compare multiple
contextual-bandit algorithms: Disjoint LinUCB, Hybrid LinUCB, Linear Thompson Sampling,
Logistic/GLM Bandits, and Neural Bandits.  The recommended first implementation is a
feature-based LinUCB/LinTS baseline with inverse-propensity-aware logging and a clear path
to hybrid and generalized-linear extensions.
\end{abstract}

\section{Scope}
The product surface is a post-purchase page in a car-rental booking service.  A user has
just completed a reservation, so the system observes contextual variables such as car type,
area, rental duration, price, device, and possibly session history.  The DSP owns a set of
ads, each with metadata such as category, target area, target car type, advertiser, budget,
and cold-start status.  The filtering layer chooses a small set of ads to pass to the
bidding layer.  The bidding layer then decides bid prices using value prediction, pacing,
and auction logic.

This document focuses on the filtering layer.  It deliberately separates filtering from
bidding because RTB contains censored feedback, budget constraints, pacing, and strategic
auction effects that are not fully captured by a standard contextual bandit.

\section{Decision Problem}
\subsection{Objects}
At interaction $t=1,2,\ldots,T$:
\begin{itemize}[leftmargin=2em]
  \item $c_t \in \mathcal{C}$ is the observed purchase context: car type, area, area type,
  rental days, price bucket, device, session-level summaries, and time features.
  \item $\mathcal{A}_t = \{a_{t1},\ldots,a_{tN_t}\}$ is the eligible ad set after hard
  business filters such as active campaign, area eligibility, creative approval, and budget availability.
  \item $z(a) \in \mathcal{Z}$ denotes ad-side features: category, target area, target car
  type, advertiser, campaign age, novelty flag, and historical priors if available.
  \item $x_t(a)=\phi(c_t,z(a))\in\R^d$ is a joint context--ad feature vector.
  \item The filtering policy selects a subset $S_t\subseteq\mathcal{A}_t$ with
  $|S_t|=K$; the bidding layer receives $S_t$.
\end{itemize}

\subsection{Reward}
Let $Y_t(a)$ be the potential outcome for ad $a$ in context $c_t$.  Depending on the
product goal, the reward can be one of:
\begin{align}
  r_t(a) &= \ind\{\text{click}\}, \\
  r_t(a) &= v_{\mathrm{cv}}\ind\{\text{conversion}\}, \\
  r_t(a) &= \alpha_{\mathrm{ctr}}\ind\{\text{click}\}
          + \alpha_{\mathrm{cvr}}v_{\mathrm{cv}}\ind\{\text{conversion}\}
          - \alpha_{\mathrm{cost}}\,\mathrm{cost}.
\end{align}
For the first filtering-layer prototype, click reward is acceptable because it is frequent
and timely.  For production optimization, the reward should move toward value-per-impression
or value-per-won-impression after accounting for delayed conversions and auction costs.

The ideal objective for top-$K$ filtering is
\begin{equation}
  \max_{\pi}\; \E\left[\sum_{t=1}^{T}\sum_{a\in S_t^\pi} r_t(a)\right],
  \quad S_t^\pi = \pi(c_t,\mathcal{A}_t),\quad |S_t^\pi|=K.
\end{equation}
Equivalently, the regret against an oracle that knows expected rewards is
\begin{equation}
  R_T = \sum_{t=1}^{T}
  \left(
    \sum_{a\in S_t^\star}\mu(c_t,a)
    - \sum_{a\in S_t^\pi}\mu(c_t,a)
  \right),
\end{equation}
where $\mu(c,a)=\E[r\mid c,a]$ and $S_t^\star$ is the top-$K$ set by $\mu$.

\subsection{Feature Design}
A linear first model requires interactions to be explicit.  A minimal feature map is:
\begin{align}
  x_t(a)=\big[&\mathrm{onehot}(\mathrm{car\_type}),
  \mathrm{onehot}(\mathrm{area}),
  \mathrm{onehot}(\mathrm{area\_type}),
  \mathrm{price\_bucket},
  \mathrm{rental\_days\_scaled},\\
  &\mathrm{onehot}(\mathrm{ad\_category}),
  \mathrm{is\_new},
  \ind\{\mathrm{context.area}=\mathrm{ad.target\_area}\},\\
  &\ind\{\mathrm{area\_type=resort}\}\times
    \ind\{\mathrm{category}\in\{\mathrm{spot},\mathrm{restaurant}\}\},\\
  &\ind\{\mathrm{car\_type}=\mathrm{ad.target\_car\_type}\}\times
    \ind\{\mathrm{category=goods}\},\\
  &\mathrm{car\_type}\times\mathrm{ad\_category},
  \mathrm{area}\times\mathrm{ad\_category}
  \big].
\end{align}
The feature map should be versioned.  Every logged decision must store the feature version,
eligible ad set, chosen set, ranking scores, and action probabilities or propensities.

\section{Feedback and Bias}
In RTB, feedback is censored.  The system observes clicks or conversions only for ads that
were selected, bid on, won, and rendered.  Therefore, naive training on won impressions can
learn the behavior of the auction and pacing system rather than intrinsic ad relevance.

A production-quality log should contain:
\begin{itemize}[leftmargin=2em]
  \item request id, context, timestamp, user/session privacy-safe keys;
  \item all eligible ads before filtering;
  \item selected top-$K$ set, scores, and propensities;
  \item bid decision, bid price, pacing multiplier, and shading multiplier;
  \item win/loss, clearing price when available, impression, click, conversion, and delayed labels;
  \item model version, feature version, and exploration parameters.
\end{itemize}

For offline evaluation, use randomized exploration traffic and estimators such as inverse
propensity scoring (IPS), self-normalized IPS, or doubly robust estimators.  Deterministic
logs from a greedy policy are insufficient for unbiased evaluation of alternative policies.

\section{Algorithm Candidates}
\subsection{Disjoint LinUCB}
Classical disjoint LinUCB maintains independent parameters per arm.  For arm $a$:
\begin{align}
  A_a &= \lambda I + \sum_{s:a_s=a} x_s x_s^\top, &
  b_a &= \sum_{s:a_s=a} r_s x_s, &
  \hat\theta_a &= A_a^{-1}b_a.
\end{align}
The score is
\begin{equation}
  p_t(a)=\hat\theta_a^\top x_t(a)+\alpha\sqrt{x_t(a)^\top A_a^{-1}x_t(a)}.
\end{equation}
It is easy to implement and interpretable, but pure per-arm learning is weak for new ads
because a newly introduced arm has no history.

\subsection{Feature-Based Shared LinUCB}
A practical first step is to represent ads by features and learn one shared model:
\begin{align}
  A &= \lambda I + \sum_s x_s x_s^\top, &
  b &= \sum_s r_s x_s, &
  \hat\theta &= A^{-1}b,\\
  p_t(a)&=\hat\theta^\top x_t(a)+\alpha\sqrt{x_t(a)^\top A^{-1}x_t(a)}.
\end{align}
This transfers learning across ads through metadata and interactions, so it is better for
cold start.  It should be the first reliable baseline for this repository.

\subsection{Hybrid LinUCB}
Hybrid LinUCB decomposes the reward into a global component and an arm-specific component:
\begin{equation}
  \E[r_t(a)] \approx z_t(a)^\top\beta + x_t(a)^\top\theta_a.
\end{equation}
Here $z_t(a)$ captures common features shared across arms, while $x_t(a)$ captures arm-local
features.  This is appropriate once there is enough data to justify advertiser- or ad-level
idiosyncratic effects without losing cold-start generalization.

\subsection{Linear Thompson Sampling}
Linear Thompson Sampling uses the same Bayesian linear-regression structure but samples a
parameter vector instead of adding an optimism bonus:
\begin{align}
  \hat\theta &= A^{-1}b, &
  \tilde\theta_t &\sim \mathcal{N}(\hat\theta, v^2 A^{-1}), &
  p_t(a)&=\tilde\theta_t^\top x_t(a).
\end{align}
It often behaves more smoothly than UCB in production ranking systems.  Because it shares
most linear-algebra machinery with LinUCB, it should be implemented as the main comparison
baseline.

\subsection{Logistic or GLM Bandits}
Click and conversion are binary, so a logistic mean function is more natural than a linear
mean:
\begin{equation}
  \Pr(r=1\mid x,a)=\sigma(\theta^\top x), \quad
  \sigma(u)=\frac{1}{1+e^{-u}}.
\end{equation}
GLM bandits are appropriate if linear predictions become poorly calibrated, especially for
low CTR regimes.  They are more complex because online confidence bounds require curvature
or approximate posterior updates.

\subsection{Neural Bandits}
Neural bandits replace the hand-engineered linear feature map with a neural representation
$f_\omega(c,a)$.  They can capture nonlinear interactions, but they require more data,
careful uncertainty estimation, and stronger monitoring.  They are not recommended as the
first production algorithm.

\section{Comparison Matrix}
\begin{longtable}{p{0.20\linewidth}p{0.24\linewidth}p{0.24\linewidth}p{0.20\linewidth}}
\toprule
Method & Strengths & Weaknesses & Recommended role \\
\midrule
Disjoint LinUCB & Simple, interpretable, per-ad parameters & Poor cold start; many ads imply many matrices & Educational baseline only \\
Shared LinUCB & Simple; fast; transfers across ads; easy cold start & Linear assumption; feature engineering required & First implementation \\
Hybrid LinUCB & Balances global transfer and arm-specific effects & More bookkeeping; more parameters & Second-stage extension \\
Linear Thompson Sampling & Smooth exploration; shares LinUCB state & Requires sampling and variance calibration & Main A/B comparison with LinUCB \\
Logistic/GLM Bandit & Better for binary rewards and calibration & More complex online inference & Use after linear baseline diagnostics \\
Neural Bandit & Nonlinear representation power & Data hungry; uncertainty is hard & Later-stage research track \\
\bottomrule
\end{longtable}

\section{Recommended Implementation Plan}
\begin{enumerate}[leftmargin=2em]
  \item Define production-grade schemas for contexts, ads, eligible sets, actions, propensities, and outcomes.
  \item Implement a shared linear model abstraction with Sherman--Morrison updates for $A^{-1}$.
  \item Implement both LinUCB and Linear Thompson Sampling on the same feature store.
  \item Add semi-bandit top-$K$ learning: update the model for each displayed ad with observed reward.
  \item Add randomized logging traffic and propensity recording before claiming offline gains.
  \item Evaluate with reward@K, CTR@K, cumulative regret in simulation, IPS/SNIPS on logs, and calibration plots.
  \item Introduce Hybrid LinUCB only after the shared model has stable logs and enough per-ad observations.
  \item Keep the bidding layer separate: expected value from the filtering layer should become one input to bid price,
  together with pacing, budget, and bid shading.
\end{enumerate}

\section{References}
\begin{thebibliography}{9}
\bibitem{li2010linucb}
L. Li, W. Chu, J. Langford, and R. E. Schapire.
\newblock A Contextual-Bandit Approach to Personalized News Article Recommendation.
\newblock WWW, 2010.

\bibitem{chu2011linear}
W. Chu, L. Li, L. Reyzin, and R. E. Schapire.
\newblock Contextual Bandits with Linear Payoff Functions.
\newblock AISTATS, 2011.

\bibitem{agrawal2013ts}
S. Agrawal and N. Goyal.
\newblock Thompson Sampling for Contextual Bandits with Linear Payoffs.
\newblock ICML, 2013.

\bibitem{li2011offline}
L. Li, W. Chu, J. Langford, and X. Wang.
\newblock Unbiased Offline Evaluation of Contextual-Bandit-based News Article Recommendation Algorithms.
\newblock WSDM, 2011.

\bibitem{wang2017rtb}
J. Wang, W. Zhang, and S. Yuan.
\newblock Display Advertising with Real-Time Bidding (RTB) and Behavioural Targeting.
\newblock Foundations and Trends in Information Retrieval, 2017.

\bibitem{zhang2014optimal}
W. Zhang, S. Yuan, and J. Wang.
\newblock Optimal Real-Time Bidding for Display Advertising.
\newblock KDD, 2014.

\bibitem{xu2015pacing}
J. Xu, K.-c. Lee, W. Li, Q. Qi, and Q. Lu.
\newblock Smart Pacing for Effective Online Ad Campaign Optimization.
\newblock KDD, 2015.

\bibitem{zhang2016censored}
W. Zhang, T. Zhou, J. Wang, and J. Xu.
\newblock Bid-aware Gradient Descent for Unbiased Learning with Censored Data in Display Advertising.
\newblock KDD, 2016.
\end{thebibliography}

\end{document}
